---
title: 'JHU - Exploratory Data Analysis, Assignment #2'
author: "jnabonne"
---

```{r global_settings, echo=F}
	library(knitr)
	opts_chunk$set(cache=F, warning=0, message=0, echo=0)
	library(dplyr)
	library(ggplot2)
```

# JHU - Exploratory Data Analysis, Assignment #2

## About
This script contains R code to generate plots asked in the [assignment #2 of eda course](https://www.coursera.org/learn/exploratory-data-analysis/peer/b5Ecl/course-project-2)

## Dataset
The dataset for this exercice available [here](https://d396qusza40orc.cloudfront.net/exdata%2Fdata%2FNEI_data.zip)

## Notes:  
* one R script per question (6 in total)
* some scripts use libraries as ggplot2 and dplyr but wont install them (IMHO, a script should not)
* scripts start by downloading and unzip dataset each time (to always get last version in case it changes)


```{r setup, echo=F, message=F, warning=F}
	# downloading dataset (each time to get its latest version)
	file_url <- "https://d396qusza40orc.cloudfront.net/exdata%2Fdata%2FNEI_data.zip"
	file_path <- "NEI_data.zip"
	download.file(file_url, destfile=file_path, method="curl")
	unzip(file_path)
	# reading (and unzipping on-the-fly) dataset's 2 files
	nei <- readRDS("summarySCC_PM25.rds")
	scc <- readRDS("Source_Classification_Code.rds")
```

---

## Q1. Have total emissions from PM2.5 decreased in US from 1999 to 2008?
```{r plot1}
	# summing emissions per years
	tot1 <- tapply(nei$Emissions, nei$year, sum)
	# building plot
	plot(names(tot1), tot1, type='b',
	     xlab="year", ylab="emissions (tons)",
	     main="Total emissions of PM2.5 in the US")
```

=> YES, globally (taking into account all type of sources), PM2.5 emissions have decreased over the 1999-2008 period

---

## Q2. Have total emissions from PM2.5 decreased in the Baltimore City?

```{r plot2}
		# filtering data for Baltimore City
	nei_baltimore <- subset(nei, nei$fips=="24510")
	# summimg emissions per years
	tot2 <- tapply(nei_baltimore$Emissions, nei_baltimore$year, sum)
	# building plot
	plot(names(tot2), tot2, type='b',
	     xlab="year", ylab="emissions (tons)",
	     main="Total emissions of PM2.5 in Baltimore City")
```

=> Over the global period, YES, emissions have decreased in Baltimore  
but we can also note that there was a clear increase between 2002 and 2005

---

# Q3. Which of 4 types of sources emission have decreased for Baltimore City?
```{r plot3}
	# summing after filtering for Baltimore and grouping by year and sources types
	tot3 <- nei %>% 
	  filter(fips=="24510") %>% 
	  group_by(year, type) %>% 
	  summarize(total=sum(Emissions))
	# building plot
	qplot(data=tot3, year, total, col=type, geom="path",
		ylab="emissions (tons)",
		main="Total emissions of PM2.5 in Baltimore City per type of sources")
```

=> POINT type of source is the only one that clearly increased during the 1999-2008 period
even if we can witness a change in the tendance since 2005.  
If NON-ROAD type registered a small increase between 2002 and 2005, 
all other types have decreased over the global period

---

# Q4. Across US, how emissions from coal combustion-related sources changed?
```{r plot 4}
	# getting idx and value of SCC codes for coal related sources
	idx_coal <- grepl("coal", scc$Short.Name, ignore.case=1)
	coal_scc <- scc[idx_coal,]$SCC	
	# filtering by coal scc codes and grouping by year
	tot4 <- nei %>% 
	  filter(SCC %in% coal_scc) %>% 
	  group_by(year) %>% 
	  summarize(total=sum(Emissions))
	# building plot
	qplot(data=tot4, year, total, geom="path", ylab="emissions (tons)",
	  main="US PM2.5 total emissions for coal combustion-related sources")
```

=> they have been decreasing except between 2002 and 2005 but have dramatically dropped since 2005

---

# Q5. How emissions from motor vehicle sources changed in Baltimore?
```{r plot 5}
	# getting idx and value of SCC codes for motor vehicule sources
	# (assuming that all vehicule source are motorized)
	idx_motor <- grepl("vehicle", scc$Short.Name, ignore.case=1)
	motor_scc <- scc[idx_motor,]$SCC
	# filtering by motor SCC codes, Baltimore and grouping by year
	tot5 <- nei %>% 
	# filter(SCC %in% motor_scc & fips=="24510" & type=="ON-ROAD") %>% 
	  filter(SCC %in% motor_scc & fips=="24510") %>% 
	  group_by(year) %>% 
	  summarize(total=sum(Emissions))
	# building plot
	qplot(data=tot5, year, total, geom="path", ylab="emissions (tons)",
	  main="PM2.5 total emissions for motor vehicules sources in Baltimore")
```

=> they have continuously decrease over the years (especially between 1999 and 2002)

---

# Q6. Compare motor vehicle emisions in Baltimore City and Los Angeles County
```{r plot 6}
	# getting idx and value of SCC codes for coal related sources
	idx_motor <- grepl("motor", scc$Short.Name, ignore.case=1)
	motor_scc <- scc[idx_motor,]$SCC	
	# filtering by motor scc codes and fips corresponding to Baltimore & LA County
	tot6 <- nei %>% 
	  filter(SCC %in% motor_scc & (fips=="24510" | fips=="06037")) %>% 
	  group_by(year, fips) %>% 
	  summarize(total=sum(Emissions))
	# building plot (but does not display it yet)
	p6 <- qplot(data=tot6, year, total, color=fips, geom="path",
	           ylab="emissions (tons)",
	           main="PM2.5 total emissions for motor vehicules sources")
	# printing plot while modifying legend on-the-fly
	print(p6 + scale_colour_hue(name="Location", breaks=c("06037", "24510"),
	                     labels=c("LA County", "Baltimore City")))
```

=> LA County sadly registers an increase in PM2.5 emissions and measures are much higher than in Baltimore  
(which has seen an increase in first years of 21th century but get back to its previous values of 1999)

---

This is the end, thanks for reading until it!